---
title: "Salmon Data Practice"
format: html
---
# Setting up packages
```{r}

library(dplyr)
library(tidyr)
library(readr)

```

# Reading data in
```{r}
catch_original <- readr::read_csv(file = "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```


```{r}
# Return the column names of my data frame
colnames(catch_original)

# Return first (or last) few lines of the data frame
head(catch_original); tail(catch_original, n = 10)

# Print a summary of each column of data
summary(catch_original)

# Return unique values in a column (in this case, the region)
unique(catch_original$Region)

# Open data frame in its own tab to see each row and column
# of the data (do in console)
View(catch_original)
```

# Selecting the columns we want

```{r}
# Select only desired columns
catch_data <- catch_original %>%
    dplyr::select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

# Check the result
head(catch_data)
```

#removing columns we don't want
```{r}
catch_data2 <- catch_original %>%
    dplyr::select(-All, -notesRegCode)
```
# more column cleaning still
```{r}
catch_data3 <- catch_original %>%
    dplyr::select(Region,
                  Year,
                  Chinook,
                  Sockeye,
                  Coho,
                  Pink,
                  Chum,
                  total = All,
                  notes = notesRegCode)

catch_data4 <- catch_original %>%
    dplyr::rename(total = All,
                  notes = notesRegCode)
```
# Changing column name to be more comprehensible
```{r}
coho_year <- catch_data %>% 
    select(Year, Oncorhynchus_kisutch = Coho)
```

# Check the structure of the data

```{r}
dplyr::glimpse(catch_data)
```

# Changing Columns using mutate

```{r}
# Make the Chinook column numeric
catch_clean <- catch_data %>%
    dplyr::mutate(Chinook = as.numeric(Chinook))
```

# Checking which value in Chinook is the problem

```{r}
(i <- which(is.na(catch_clean$Chinook)))
```

# Looking at the problem row only
```{r}
catch_data[i, ]
```

# Fixing the problem in the row with the if_else func

```{r}
# Use a conditional to fix the bad value, then make the column numeric
catch_clean <- catch_data %>%
    dplyr::mutate(
        Chinook = dplyr::if_else(condition = Chinook == "I", 
            true = "1",
            false = Chinook),
        Chinook = as.numeric(Chinook))

# Check the result
catch_clean[i, ]
```

# Using pivot to wrangle data into "tidy" format

```{r}
# Pivot all columns except Region and Year into long format
catch_long <- catch_clean %>% 
    tidyr::pivot_longer(cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch_thousands")

# Check the result
head(catch_long)
```

# Opposite of the pivot_longer is pivot_wider, changing back to pivot_wider

```{r}
# Pivot the data back into wide format
catch_wide <- catch_long %>%
    tidyr::pivot_wider(names_from = species,
        values_from = catch_thousands,
        values_fill = 0)

# Check the result
head(catch_wide)
```

# to check how to use either type ?tidyr::pivot_wider into the Consule

# Adding a new column- with units of fish instead of thousands- with mutate

```{r}
# Convert to actual fish numbers, then remove the superseded column
catch_long <- catch_long %>%
    dplyr::mutate(catch = catch_thousands * 1000) %>% 
    dplyr::select(-catch_thousands)

# Check the result
dplyr::glimpse(catch_long)
```

# Summary statistics with dpylr!

```{r}
unique(catch_long$Region)
```

# we are spliting the table into regions, calculating the mean for each region, and then consolidating the table again.

```{r}
# Group by region, then calculate averages within the established groups
mean_region <- catch_long %>%
    dplyr::group_by(Region) %>%
    dplyr::summarize(mean_catch = mean(catch, na.rm = TRUE),
        .groups = "drop")

# Check the result
dplyr::glimpse(mean_region)
```

# Example of another .groupby use

```{r}
# Count rows per region
n_region <- catch_long %>%
    dplyr::group_by(Region) %>%
    dplyr::summarize(count = dplyr::n(), 
        .groups = "drop")

# Check the result
head(n_region)
```

# Conditionally filtering groups using filter()

```{r}
# Limit catch to the SSE region
sse_catch <- catch_long %>%
    dplyr::filter(Region == "SSE")

# Check the result
unique(sse_catch$Region) 
```

# Sort data with arrange ()

```{r}
# Calculate mean catch by region and sort by catch
mean_region <- catch_long %>%
    dplyr::group_by(Region) %>%
    dplyr::summarize(mean_catch = mean(catch, na.rm = TRUE),
        .groups = "drop") %>%
    dplyr::arrange(mean_catch)

# Check the result
head(mean_region)
```

# sort in ascendending order

```{r}
# Calculate mean catch by region and sort *descending* by catch
mean_region <- catch_long %>%
    dplyr::group_by(Region) %>%
    dplyr::summarize(mean_catch = mean(catch, na.rm = TRUE),
        .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(mean_catch))

# Check the result
head(mean_region)
```

# Separate columns with separate_wider_delim() func
## make a new data frame to demonstrate func
```{r}
sites_df <- data.frame(site = c("HAW-101", "HAW-103", "OAH-320", "OAH-219", "MAU-039"))

# Check that out
head(sites_df)
```

```{r}
# Split site by the dash between the site letters and its code
sites_df %>%
    tidyr::separate_wider_delim(cols = site, delim = "-",
        names = c("island", "site_number"), cols_remove = FALSE)
```

# Uniting columns with unite()
## making new dataframe to demonstrate this
```{r}
dates_df <- data.frame(
    year  = rep(x = "1930", times = 3),
    month = rep(x = "12",   times = 3),
    day   = 14:16)

# Check that
dates_df
```

```{r}
# Combine those columns into a single 'date' column
dates_df %>%
    tidyr::unite(col = date, 
        year, month, day,
        sep = "-", 
        remove = FALSE)
```
# Putting it all together! doing everything we did but more concisely in one bloc

```{r}
# Grab the data
catch_original <- readr::read_csv(file = "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1",
    show_col_types = FALSE)

# Drop unwanted columns, fix "I" typo, reshape to long, transform catch
# into individual fish count, calculate mean catch by region, and sort
# in descending order
mean_region <- catch_original %>%
    dplyr::select(-All, -notesRegCode) %>% 
    dplyr::mutate(
        Chinook = dplyr::if_else(Chinook == "I", 
            true = "1", false = Chinook),
        Chinook = as.numeric(Chinook)) %>% 
    tidyr::pivot_longer(-c(Region, Year), 
        names_to = "species", 
        values_to = "catch") %>%
    dplyr::mutate(catch = catch*1000) %>% 
    dplyr::group_by(Region) %>% 
    dplyr::summarize(mean_catch = mean(catch, na.rm = TRUE),
        .groups = "drop") %>% 
    dplyr::arrange(dplyr::desc(mean_catch))

# Check the result
head(mean_region)
```

